---
layout: post
title: Mixed Effects to the Rescue!
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lme4)
library(data.table)
```



Linear and logistic regression work great on canned example data found on blogs and websites but
run into many problems when they are deployed in the wild.  Data is often grouped into categories 
that are imbalanced (some customers have more purchasing history than others), hierarchical 
(products can be part of one 'family' with slight differences), and non independent (sales people 
have different impacts on outcomes).  In each of these examples, standard regression 
techniques fail to properly address the structure of the data.  

## Linear Mixed Models - Regression for Real World Data
Linear Mixed Models (LMM) work like 
standard regression, but with additional terms called random effects that capture variation not 
explained by the independent variables used to create the model. 


#### THE DATA

For this post I'm going to use some fast food data provided by tidy tuesday.  If you haven't heard of
[Tidy Tuesday](https://twitter.com/hashtag/tidytuesday?lang=en) - check it out, lots of great visualization examples in R.

This data works great for LMM's because its grouped, it has inherent hierarches, and there's large 
imbalances of items between groups.  All of these characteristcs would be a headache for traditional 
linear models.  We'll see how adding random effects helps. 

```{r}
fast_food = fread('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-09-04/fastfood_calories.csv')
fast_food[,item := tolower(item)]
fast_food[,has_chicken := grepl('chicken', item)]
fast_food[,chicken_strip := grepl('strip|popcorn|nuggets|chicken fries|tenders',item)]
fast_food[,is_salad := grepl('salad',item)]

subway_subset = fast_food[restaurant == 'Subway']
fast_food = fast_food[!(restaurant == 'Subway' & grepl('footlong',item)) & !(restaurant == 'Subway' & grepl('kids mini',item))]
fast_food[restaurant == 'Subway', sandwich := grepl('6', item)]
fast_food[, is_wrap := grepl('wrap', item)]
fast_food[is.na(sandwich),sandwich := grepl('sandwich', item)]

```

## Difference Between Random and Fixed Effects

Fixed Effects are categorical predictors where you know all the possible levels.  Some Examples:

- Drink Size (S/M/L)

- Is the chicken fried? (Y/N)

Random Effects are categorical predictors that your data only captures a subset of, of all possible 
levels in the full data.  In practice, they allow differing levels of variation and association 
to the response creating non constant error distributions.

- Restaurant Chain 

- Items on customer orders

## Benefits of incorporating random effects in your model

Random effects have several benefits, the main ones are:

- Enable us to make better predictions on sparse groups using partial pooling

- Allow us to explicitly model non independence in data, such as repeated measurements

- Better capture hierarchical or clustered data such as geography or product lines


### Imbalanced Data
The Random Effect term in a LMM uses a property called *Partial Pooling*.  Partial pooling allows the model
to create an overall effect for a predictor that will pull smaller groups toward the group effect while 
larger groups will be more distinct.  This is similar to how in Bayesian statistics repeated measurements
improve your confidence in an estimate.

In the below example, we try to predict the amount of calories in a salad based on the amount of fat.
By incorporating a random effect for the restaurant, we can get a stronger estimate for restaurants
with fewer salad options.

```{r}
salad_data = fast_food[is_salad == T]
salad_data[,fried := grepl('crispy',item)]
mixed_effects = lmer(calories ~ total_fat + (1|restaurant), data = salad_data)
fixed_effects = lm(calories ~ total_fat, data = salad_data)
AIC(fixed_effects, mixed_effects)
```

### Non Independence 

Non independence of measurements is very common in real world data.  Below is an example using 
chicken strips, nuggets, fries, and tenders.  Each restaurant has a signature method for preparing 
chicken and this preparation will impact the caloric content.  Treating each restaurant as a fixed effect
rapidly increases the degrees of freedom - and doesn't consider that this data only represents a subset
of all restaurants with chicken tenders!

Making restaurant a random effect better represents the data and makes the model more generalizeable.
The mixed effect model also has a lower Akaike's Information Criterion (a measure of how model fit - lower 
better).



```{r}
chicken_strip_data = fast_food[chicken_strip == T]
chicken_strip_data[,count := as.integer(sub('[^0-9]+','',item))]
chicken_strip_data = chicken_strip_data[!is.na(count)]
mixed_effects_chicken = lmer(calories ~ count + (1|restaurant), data = chicken_strip_data)
fixed_effects_chicken = lm(calories ~ count, data = chicken_strip_data)
AIC(mixed_effects_chicken, fixed_effects_chicken)
```

### Hierarchical Relationships

Nested relationships are also very common in day to day data.  In the example below - subway item 
belongs to a class of item , sandwich, salad, etc and a flavor meatball, tuna, etc.  Representing this
structure in the data would require many separate models if using a traditional fixed effects model.
By incorporating random effects, we can use the structure in the data to get a better overall fit.

```{r}
subway_subset[,item_general := ifelse(grepl('6""', item),'6_sandwich',NA)]
subway_subset[is.na(item_general),item_general := ifelse(grepl('footlong', item),'12_sandwich',NA)]
subway_subset[is.na(item_general),item_general := ifelse(grepl('kids mini', item),'3_sandwich',NA)]


subway_subset[is.na(item_general),item_general := ifelse(grepl('salad', item),'salad',NA)]
subway_subset[is.na(item_general),item_general := ifelse(grepl('wrap', item),'wrap',NA)]
subway_subset[is.na(item_general),item_general := ifelse(grepl('pizza', item),'pizza',NA)]

subway_subset[,item := sub('6""|footlong|kids mini','',item)]
subway_subset[,item := sub('6""|footlong|kids mini|salad|wrap|pizza','',item)]

subway_model = lmer(protein ~ total_carb +  (1|item_general) + (1|item), data = subway_subset)
```


## sauce

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5970551/

http://lme4.r-forge.r-project.org/book/Ch2.pdf

https://stats.idre.ucla.edu/other/mult-pkg/introduction-to-generalized-linear-mixed-models/

https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf

